Weighted Model Counting Without Parameter Variables

### 2021-02-21 doc/paper3/paper.tex
Some writing

@@

\begin{definition} \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions. Operations
  such as addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y), \quad \text{and} \quad (fg)(Y) \coloneqq
    f(Y)g(Y)
  \]
  for all $Y \subseteq X$. This means that binary operations on pseudo-Boolean
  functions inherit properties such as associativity and commutativity.
\end{definition}

\begin{definition} \label{def:restriction}
  Let $f\colon 2^X \to \mathbb{R}$ be a pseudo-Boolean function, and $x \in X$.
  Then $f|_{x=0}, f|_{x=1}\colon 2^X \to \mathbb{R}$ are \emph{restrictions} of
  $f$ defined as
  \[
    f|_{x=0}(Y) \coloneqq f(Y \setminus \{x\}), \quad \text{and} \quad
    f|_{x=1}(Y) \coloneqq f(Y \cup \{x\})
  \]
  for all $Y \subseteq X$.
\end{definition}

\begin{definition} \label{def:projection}
  Let $X$ be a set. For any $x \in X$, \emph{projection} $\exists_x$ is an
  endomorphism $\exists_x\colon \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ defined
  as
  \[
    \exists_xf \coloneqq f|_{x=1} + f|_{x=0}
  \]
  for any $f\colon 2^X \to \mathbb{R}$.
\end{definition}

% NOTE: ...that we will need.
% NOTE: The proof immediately comes from the definition (of operations).
\begin{proposition}[Basic properties] \label{prop:basic}
  For any propositional formulas $\phi$ and $\psi$, and $a, b, c, d \in
  \mathbb{R}$,
  \begin{itemize}
  \item $[\phi]^a_b = [\neg \phi]^b_a$;
  \item $c + [\phi]^a_b = [\phi]^{a+c}_{b+c}$;
  \item $c \cdot [\phi]^a_b = [\phi]^{ac}_{bc}$;
  \item $[\phi]^a_b \cdot [\phi]^c_d = [\phi]^{ac}_{bd}$;
  \item $[\phi]^1_0 \cdot [\psi]_0^1 = [\phi \land \psi]_0^1$.
  \end{itemize}
  And for any pair of pseudo-Boolean functions $f, g \colon 2^X \to \mathbb{R}$
  and $x \in  X$, $(fg)|_{x=i} = f|_{x=i} \cdot g|_{x=i}$ for $i = 0, 1$.
\end{proposition}

\begin{definition}[old WMC instance]
  An \emph{old WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of indicator variables, $X_P$ is the set of parameter variables (with
  $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula in CNF over
  $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
  X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of the
  instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition} % TODO: add a citation for this.

\begin{remark}
  Encodings such as \texttt{cd05} and \texttt{cd06} are \emph{not} WMC
  encodings. Instead, they encode Bayesian network inference into instances of
  the \emph{minimum-cardinality} WMC problem, where the answer is defined to be
  $\sum_{Y \models \phi\text{, }|Y| = k} \prod_{Y \models l}
  w(l)$, where $k = \min_{Y \models \phi\text{, }Y \ne \emptyset}
  |Y|$, if $k$ exists, otherwise the answer is zero. This additional
  condition on model cardinality becomes necessary because these encodings
  eliminate clauses of the form $p \Rightarrow i$, where $p \in X_P$ is a
  parameter variable, and $i \in X_I$ is an indicator variable. Nonetheless, our
  transformation algorithm still works on such encodings, although the
  experimental results are discouraging because they use approximately twice as
  many indicator variables. For instance, each binary variable of a Bayesian
  network is encoded using two indicator variables while one would suffice.
\end{remark} % TODO: perhaps add this claim as a corollary

\begin{definition}[new WMC instance] \label{def:new_wmc}
  A \emph{new WMC instance} is a tuple $(F, X, \omega)$, where $X$ is the set of
  variables, $F$ is a set of pseudo-Boolean functions $2^X \to \mathbb{R}$, and
  $\omega \in \mathbb{R}$ is the scaling factor. The \emph{answer} of the
  instance is $\omega \cdot \left(\exists_X\prod_{f \in F}f\right)(\emptyset)$.
\end{definition}

@@

\paragraph{Notes.}
\begin{itemize}
\item Apparently, the \textsc{DPMC} paper already shows that taking the first
  offered decomposition tree is best.
\item It is already well-known that WMC is FPT.
\item Weights on literals other than the positive literals that correspond to
  variables in $X_P$ are redundant as they either are equal to one or duplicate
  an already-defined weight.
\item Mention that formulas, clauses, and models are all treated as sets.
\item The constant is inspired by the \texttt{bklm16} encoding.
\item We don't make a distinction between a real number and a (pseudo-Boolean)
  function that always returns that number.
\item Throughout the paper, projection is assumed to have the lowest precedence
  (e.g., $\exists_x fg = \exists_x (fg)$ and not $(\exists_xf)g$).
\end{itemize}

@@

\begin{algorithm}
  \caption{WMC instance transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{an (old-format) WMC instance $(\phi, X_I, X_P, w)$}
  \KwResult{a (new-format) WMC instance $(F, X, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
  replace every variable $v$ in $F$ with $\rename{$v$}$\;
  \Return{$(F, X_I, \omega)$}\;
  \Fn{\rename{$v$}}{
    $S \gets \{u \in X_P \mid u \le v\}$\;
    \lIf{$S = \emptyset$}{\Return{$v$}}
    \Return{$v - d(\max S)$}\;
  }
\end{algorithm}

\subsection{Correctness}

@@

\begin{lemma} \label{lemma:sum}
  For any pseudo-Boolean function $f\colon 2^X \to \mathbb{R}$, $(\exists_X
  f)(\emptyset) = \sum_{Y \subseteq X} f(Y)$.
\end{lemma}
\begin{proof}
  For base case, let $X = \{x\}$. Then
  \[
    \exists_xf = f|_{x=1} + f|_{x=0},
  \]
  by \cref{def:projection} and so
  \[
    (\exists_xf)(\emptyset) = f|_{x=1}(\emptyset) + f|_{x=0}(\emptyset)
  \]
  by \cref{def:operations}. Then
  \[
    f|_{x=1}(\emptyset) = f(\{x\}), \quad \text{and} \quad f|_{x=0}(\emptyset) =
    f(\emptyset)
  \]
  by \cref{def:restriction}. It follows that
  \[
    (\exists_xf)(\emptyset) = f(\{x\}) + f(\emptyset) = \sum_{Y \subseteq \{x\}}
    f(Y)
  \]
  as required. This easily extends to $|X| > 1$ by the definition of projection
  on sets of variables. % TODO: once I've defined it, add a reference here.
\end{proof}

% NOTE: mention that this is similar (but not equal) to the DPMC/ADDMC result
\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be an old WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a new WMC instance with the same answer.
\end{proposition}
\begin{proof}
  The answer of \cref{eq:new_wmc} is
  \begin{align*}
    \left(\exists_{X_I \cup X_P} \prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset) &= \sum_{Y \subseteq X_I \cup X_P} \left(\prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)}\right)(Y) \\
&= \sum_{Y \subseteq X_I \cup X_P} \underbrace{\left(\prod_{c \in \phi} [c]_0^1\right)}_f(Y) \underbrace{\left(\prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)}\right)}_g(Y)
  \end{align*}
  by \cref{def:new_wmc,lemma:sum,def:operations}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

\begin{theorem}[Correctness]
  \Cref{alg:transformation}, when given an old WMC instance $(\phi, X_I, X_P,
  w)$, returns a new WMC instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}
\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc}
  \end{equation}
  is a new WMC instance with the same answer as the given old WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \prod_{c \in \phi} [c]_0^1 \prod_{x \in X_I
        \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \cref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

  Let us first consider how the unfinished WMC instance $(F, X_I, \omega)$ after
  the loop on \crefrange{line:foreach1start}{line:foreach1end} differs from
  \cref{eq:new_wmc}. Note that \cref{alg:transformation} leaves each $c \in
  \phi_*$ unchanged, i.e., adds $[c]_0^1$ to $F$. We can then fix an arbitrary
  $p \in X_P$ and let $F_p$ be the set of functions added to $F$ as a
  replacement of $\phi_p$. It is sufficient to show that
  \begin{equation} \label{eq:to_show}
    \omega \prod_{f \in F_p} f = \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1.
  \end{equation}
  Note that under \cref{condition:d02},
  \[
    \bigwedge_{c \in \phi_p} c \equiv p \Leftrightarrow \bigwedge_{i=1}^n l_i
  \]
  for some family of indicator variable literals $(l_i)_{i=1}^n$. Thus,
  \[
    \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 = \exists_p
    [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1.
  \]
  If $w(p) = 1$, then
  \begin{equation} \label{eq:bigsums}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = \exists_p \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
      l_i \right]_0^1\right|_{p=0}
  \end{equation}
  by \cref{def:projection}. Since for any input, $\bigwedge_{i=1}^n l_i$ is
  either true or false, exactly one of the two summands in \cref{eq:bigsums}
  will be equal to one, and the other will be equal to zero, and so
  \[
    \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
      \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
        l_i \right]_0^1\right|_{p=0} = 1,
  \]
  where $1$ is a pseudo-Boolean function that always returns one. On the other
  side of \cref{eq:to_show}, since $F_p = \emptyset$, and $\omega$ is unchanged,
  we get $\omega\prod_{f \in F_p} f = 1$, and so \cref{eq:to_show} is satisfied
  under \cref{condition:d02} when $w(p) = 1$.

>>>>

\section{Pseudo-Boolean Functions}

@@

\begin{definition}[Operations] \label{def:operations}
  Let $f, g\colon 2^X \to \mathbb{R}$ be pseudo-Boolean functions, $x, y \in X$,
  $Y = \{y_i\}_{i=1}^n \subseteq X$, and $r \in \mathbb{R}$. Operations such as
  addition and multiplication are defined pointwise as
  \[
    (f+g)(Y) \coloneqq f(Y)+g(Y), \quad \text{and} \quad (f \cdot g)(Y)
    \coloneqq f(Y) \cdot g(Y).
  \]
  Note that this means that binary operations on pseudo-Boolean functions
  inherit properties such as associativity and commutativity. By not
  distinguishing between a real number and a pseudo-Boolean function that always
  returns that number, we can use the same definitions to define \emph{scalar}
  operations as
  \[
    (r+f)(Y) = r+f(Y), \quad \text{and} \quad (r \cdot f)(Y) = r \cdot f(Y).
  \]

  \emph{Restrictions} $f|_{x=0}, f|_{x=1}\colon 2^X \to \mathbb{R}$ of $f$ are
  defined as
  \[
    f|_{x=0}(Y) \coloneqq f(Y \setminus \{x\}), \quad \text{and} \quad
    f|_{x=1}(Y) \coloneqq f(Y \cup \{x\})
  \]
  for all $Y \subseteq X$.

  \emph{Projection} $\exists_x$ is an endomorphism $\exists_x\colon
  \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ defined as
  \[
    \exists_xf \coloneqq f|_{x=1} + f|_{x=0}.
  \]
  Since projection is commutative (i.e., $\exists_x\exists_yf =
  \exists_y\exists_xf$) \cite{DBLP:conf/aaai/DudekPV20,DBLP:conf/cp/DudekPV20},
  we can define $\exists_Y\colon \mathbb{R}^{2^X} \to \mathbb{R}^{2^X}$ as
  $\exists_Y \coloneqq \exists_{y_1}\exists_{y_2}\dots\exists_{y_n}$. Throughout
  the paper, projection is assumed to have the lowest precedence (e.g.,
  $\exists_x fg = \exists_x (fg)$).
\end{definition}

% NOTE: ...that we will need.
% NOTE: The proof immediately comes from the definition (of operations).
\begin{proposition}[Basic Properties] \label{prop:basic}
  For any propositional formulas $\phi$ and $\psi$, and $a, b, c, d \in
  \mathbb{R}$,
  \begin{itemize}
  \item $[\phi]^a_b = [\neg \phi]^b_a$;
  \item $c + [\phi]^a_b = [\phi]^{a+c}_{b+c}$;
  \item $c \cdot [\phi]^a_b = [\phi]^{ac}_{bc}$;
  \item $[\phi]^a_b \cdot [\phi]^c_d = [\phi]^{ac}_{bd}$;
  \item $[\phi]^1_0 \cdot [\psi]_0^1 = [\phi \land \psi]_0^1$.
  \end{itemize}
  And for any pair of pseudo-Boolean functions $f, g \colon 2^X \to \mathbb{R}$
  and $x \in  X$, $(fg)|_{x=i} = f|_{x=i} \cdot g|_{x=i}$ for $i = 0, 1$.
\end{proposition}

@@

\paragraph{TODO: Mention}
\begin{itemize}
\item Weights on literals other than the positive literals that correspond to
  variables in $X_P$ are redundant as they either are equal to one or duplicate
  an already-defined weight.
\item Mention that formulas, clauses, and models are all treated as sets.
\end{itemize}

\section{Weighted Model Counting}

\begin{definition}[WMC Instance \cite{DBLP:journals/ai/ChaviraD08} (TODO: not
  exactly though)]
  A \emph{WMC instance} is a tuple $(\phi, X_I, X_P, w)$, where $X_I$ is
  the set of indicator variables, $X_P$ is the set of parameter variables (with
  $X_I \cap X_P = \emptyset$), $\phi$ is a propositional formula in CNF over
  $X_I \cup X_P$, and $w\colon X_I \cup X_P \cup \{\neg x \mid x \in X_I \cup
  X_P\} \to \mathbb{R}$ is the weight function. The \emph{answer} of the
  instance is $\sum_{Y \models \phi} \prod_{Y \models l} w(l)$.
\end{definition}

\begin{remark}
  Encodings such as \texttt{cd05} and \texttt{cd06} are \emph{not} WMC
  encodings. Instead, they encode Bayesian network inference into instances of
  the \emph{minimum-cardinality} WMC problem, where the answer is defined to be
  $\sum_{Y \models \phi\text{, }|Y| = k} \prod_{Y \models l}
  w(l)$, where $k = \min_{Y \models \phi\text{, }Y \ne \emptyset}
  |Y|$, if $k$ exists, otherwise the answer is zero. This additional
  condition on model cardinality becomes necessary because these encodings
  eliminate clauses of the form $p \Rightarrow i$, where $p \in X_P$ is a
  parameter variable, and $i \in X_I$ is an indicator variable. Nonetheless, our
  transformation algorithm still works on such encodings, although the
  experimental results are discouraging because they use approximately twice as
  many indicator variables. For instance, each binary variable of a Bayesian
  network is encoded using two indicator variables while one would suffice.
\end{remark}

\begin{definition}[PBP Instance] \label{def:new_wmc}
  A \emph{pseudo-Boolean projection} (PBP) instance is a tuple $(F, X, \omega)$,
  where $X$ is the set of variables, $F$ is a set of pseudo-Boolean functions
  $2^X \to \mathbb{R}$, and $\omega \in \mathbb{R}$ is the scaling factor. The
  \emph{answer} of the instance is $\omega \cdot \left(\exists_X\prod_{f \in
      F}f\right)(\emptyset)$.
\end{definition}

NOTE: The constant is inspired by the \texttt{bklm16}
\cite{DBLP:conf/ecai/BartKLM16} encoding.

@@

\begin{algorithm}[t]
  \caption{WMC instance transformation}
  \label{alg:transformation}
  \SetKwFunction{rename}{rename}
  \SetKwProg{Fn}{Function}{:}{}
  \KwData{WMC instance $(\phi, X_I, X_P, w)$}
  \KwResult{PBP instance $(F, X, \omega)$}
  $F \gets \emptyset$\;
  $\omega \gets 1$\;
  let $d\colon X_P \to \mathbb{N}$ be defined as $p \mapsto |\{ o \in X_P \mid o
  \le p \}|$\;
  \ForEach{clause $c \in \phi$\label{line:foreach1start}}{
    \uIf{$c \cap X_P = \{ p \}$ for some $p$ \textnormal{\textbf{and}} $w(p) \ne
      1$}{
      \uIf{$|c| = 1$}{$\omega \gets \omega \times w(p)$\;}
      \Else{
        $F \gets F \cup \left\{ \left[ \bigwedge_{l \in c \setminus
              \{p\}} \neg l \right]^{w(p)}_1 \right\}$\;
      }
    }
    \ElseIf{$\{p \mid \neg p \in c\} \cap X_P  = \emptyset$}{
      $F \gets F \cup \{ [c]^1_0 \}$\; \label{line:foreach1end}
    }
  }
  \ForEach{indicator variable $v \in X_I$ \label{line:foreach2start}}{
    \If{$\{[v]_1^p, [\neg v]_1^q\} \subseteq F$ for some $p$ and $q$}{
      $F \gets F \setminus \{ [v]_1^p, [\neg v]_1^q \} \cup \{ [v]_q^p
      \}$\; \label{line:foreach2end}
    }
  }
  replace every variable $v$ in $F$ with $\rename{$v$}$\;
  \Return{$(F, X_I, \omega)$}\;
  \Fn{\rename{$v$}}{
    $S \gets \{u \in X_P \mid u \le v\}$\;
    \lIf{$S = \emptyset$}{\Return{$v$}}
    \Return{$v - d(\max S)$}\;
  }
\end{algorithm}

\subsection{Proof of Correctness}

@@

\begin{lemma} \label{lemma:sum}
  For any pseudo-Boolean function $f\colon 2^X \to \mathbb{R}$,
  \[
    (\exists_X f)(\emptyset) = \sum_{Y \subseteq X} f(Y).
  \]
\end{lemma}
\begin{proof}
  For base case, let $X = \{x\}$. Then
  \[
    (\exists_xf)(\emptyset) = (f|_{x=1} + f|_{x=0})(\emptyset) =
    f|_{x=1}(\emptyset) + f|_{x=0}(\emptyset) = f(\{x\}) + f(\emptyset) =
    \sum_{Y \subseteq \{x\}} f(Y).
  \]
  This easily extends to $|X| > 1$ by the definition of projection on sets of
  variables.
\end{proof}

% NOTE: mention that this is similar (but not equal) to the DPMC/ADDMC result
\begin{proposition} \label{prop:equivalence}
  Let $(\phi, X_I, X_P, w)$ be a WMC instance. Then
  \begin{equation}
  \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
        x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
    1\right) \label{eq:new_wmc}
  \end{equation}
  is a PBP instance with the same answer.
\end{proposition}
\begin{proof}
  The answer of \cref{eq:new_wmc} is
  \begin{align*}
    \left(\exists_{X_I \cup X_P} \left(\prod_{c \in \phi} [c]_0^1\right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset) &= \sum_{Y \subseteq X_I \cup X_P} \left(\left(\prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)}\right)(Y) \\
&= \sum_{Y \subseteq X_I \cup X_P} \underbrace{\left(\prod_{c \in \phi} [c]_0^1\right)}_f(Y) \underbrace{\left(\prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)}\right)}_g(Y)
  \end{align*}
  by \cref{lemma:sum}. Note that
  \[
    f(Y) =
    \begin{cases}
      1 & \text{if } Y \models \phi, \\
      0 & \text{otherwise},
    \end{cases}
    \quad
    \text{and}
    \quad
    g(Y) = \prod_{Y \models l} w(l),
  \]
  which means that
  \[
    \sum_{Y \subseteq X_I \cup X_P} f(Y)g(Y) = \sum_{Y \models \phi} \prod_{Y
      \models l} w(l)
  \]
  as required.
\end{proof}

\begin{theorem}[Correctness]
  \Cref{alg:transformation}, when given a WMC instance $(\phi, X_I, X_P,
  w)$, returns PBP instance with the same answer, provided the following
  conditions are satisfied:
  \begin{enumerate}
  \item for all indicator variables $i \in X_I$, $w(i) = w(\neg i) = 1$,
  \item and either
    \begin{enumerate}
    \item for all parameter variables $p \in X_P$, there is a non-empty family
      of literals $(l_i)_{i=1}^n$ such that \label{condition:d02}
      \begin{enumerate}
      \item $w(\neg p) = 1$,
      \item $l_i \in X_I$ or $\neg l_i \in X_I$ for all $i = 1, \dots, n$,
        \label{condition:equivalence1}
      \item and $\{ c \in \phi \mid p \in c \text{ or } \neg p \in c \} =
        \left\{p \lor \bigvee_{i=1}^n \neg l_i \right\} \cup \{ l_i \lor \neg p
        \mid i = 1, \dots, n \}$; \label{condition:equivalence2}
      \end{enumerate}
    \item or for all parameter variables $p \in X_P$, \label{condition:bklm16}
      \begin{enumerate}
      \item $w(p) + w(\neg p) = 1$,
      \item for any clause $c \in \phi$, $|c \cap X_P| \le 1$,
      \item there is no clause $c \in \phi$ such that $\neg p \in c$,
      \item if $\{ p \} \in \phi$, then there is no clause $c \in \phi$ such
        that $c \ne \{ p \}$ and $p \in c$, \label{cond:just_parameter}
        % 'prime implicants are logically disjoint'
      \item and for any $c, d \in \phi$ such that $c \ne d$, $p \in c$ and $p
        \in d$, $\bigwedge_{l \in c \setminus \{ p \}} \neg l \land \bigwedge_{l
          \in d \setminus \{ p \}} \neg l$ is false. \label{cond:disjoint}
      \end{enumerate}
    \end{enumerate}
  \end{enumerate}
\end{theorem}
\begin{proof}
  By \cref{prop:equivalence},
  \begin{equation}
    \left(\left\{[c]_0^1 \;\middle|\; c \in \phi\right\} \cup \left\{[x]_{w(\neg
          x)}^{w(x)} \;\middle|\; x \in X_I \cup X_P\right\}, X_I \cup X_P,
      1\right) \label{eq:new_wmc2}
  \end{equation}
  is a PBP instance with the same answer as the given WMC instance. By
  \cref{def:new_wmc}, its answer is
  \begin{equation} \label{eq:answer}
    \left(\exists_{X_I \cup X_P} \left( \prod_{c \in \phi} [c]_0^1 \right) \prod_{x \in X_I \cup X_P} [x]_{w(\neg x)}^{w(x)} \right)(\emptyset)
  \end{equation}
  Since both \cref{condition:d02,condition:bklm16} ensure that each clause in
  $\phi$ has at most one parameter variable, we can partition $\phi$ into
  $\phi_* \coloneqq \{c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \emptyset \}$
  and $\phi_p \coloneqq \{ c \in \phi \mid \mathtt{Vars}(c) \cap X_P = \{ p \}
  \}$ for all $p \in X_P$. We can then use \cref{thm:early} to reorder
  \eqref{eq:answer} into
  \[
    \left(\exists_{X_I} \left( \prod_{x \in X_I} [x]_{w(\neg x)}^{w(x)} \right)
      \left( \prod_{c \in \phi_*} [c]_0^1 \right) \prod_{p \in X_P} \exists_p
      [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 \right)(\emptyset).
  \]

  Let us first consider how the unfinished WMC instance $(F, X_I, \omega)$ after
  the loop on \crefrange{line:foreach1start}{line:foreach1end} differs from
  \eqref{eq:new_wmc2}. Note that \cref{alg:transformation} leaves each $c \in
  \phi_*$ unchanged, i.e., adds $[c]_0^1$ to $F$. We can then fix an arbitrary
  $p \in X_P$ and let $F_p$ be the set of functions added to $F$ as a
  replacement of $\phi_p$. It is sufficient to show that
  \begin{equation} \label{eq:to_show}
    \omega \prod_{f \in F_p} f = \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1.
  \end{equation}
  Note that under \cref{condition:d02},
  \[
    \bigwedge_{c \in \phi_p} c \equiv p \Leftrightarrow \bigwedge_{i=1}^n l_i
  \]
  for some family of indicator variable literals $(l_i)_{i=1}^n$. Thus,
  \[
    \exists_p [p]_{w(\neg p)}^{w(p)} \prod_{c \in \phi_p} [c]_0^1 = \exists_p
    [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1.
  \]
  If $w(p) = 1$, then
  \begin{equation} \label{eq:bigsums}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = \exists_p \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1 = \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
    \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
      l_i \right]_0^1\right|_{p=0}.
  \end{equation}
  Since for any input, $\bigwedge_{i=1}^n l_i$ is either true or false, exactly
  one of the two summands in \cref{eq:bigsums} will be equal to one, and the
  other will be equal to zero, and so
  \[
    \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i
      \right]_0^1\right|_{p=1} + \left.\left[ p \Leftrightarrow \bigwedge_{i=1}^n
        l_i \right]_0^1\right|_{p=0} = 1,
  \]
  where $1$ is a pseudo-Boolean function that always returns one. On the other
  side of \cref{eq:to_show}, since $F_p = \emptyset$, and $\omega$ is unchanged,
  we get $\omega\prod_{f \in F_p} f = 1$, and so \cref{eq:to_show} is satisfied
  under \cref{condition:d02} when $w(p) = 1$.

  If $w(p) \ne 1$, then
  \[
    F_p = \left\{ \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} \right\},
  \]
  and $\omega = 1$, and so we want to show that
  \[
    \left[ \bigwedge_{i = 1}^n l_i \right]_1^{w(p)} = \exists_p [p]_1^{w(p)}
    \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1,
  \]
  and indeed
  \begin{align*}
    \exists_p [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 &= \left.\left( [p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 \right)\right|_{p=1} + \left.\left([p]_1^{w(p)} \left[ p \Leftrightarrow \bigwedge_{i=1}^n l_i \right]_0^1 \right)\right|_{p=0} \\
                                                                                      &= w(p) \cdot \left[ \bigwedge_{i=1}^n l_i \right]_0^1 + \left[ \bigwedge_{i=1}^n l_i \right]_1^0 = \left[ \bigwedge_{i=1}^n l_i \right]_1^{w(p)}.
  \end{align*}
  This finishes the proof of the correctness of the first `foreach' loop under
  \cref{condition:d02}.

  Now let us assume \cref{condition:bklm16}. We still want to prove
  \cref{eq:to_show}. If $w(p) = 1$, then $F_p = \emptyset$, and $\omega = 1$,
  and so the left-hand side of \cref{eq:to_show} is equal to one. Then the
  right-hand side is
  \[
    \exists_p [p]_0^1 \prod_{c \in \phi_p} [c]_0^1 = \exists_p \left[ p \land
      \bigwedge_{c \in \phi_p} c \right]_0^1 = \exists_p [p]_0^1 = 0 + 1 = 1
  \]
  since $p \in c$ for every clause $c \in \phi_p$.
